---
title: "Quickly get time series data ready for analysis with padr"
author: "Edwn Thoen"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Working with dates and times

Raw data containing date or time variables is often not directly ready for analysis. Data might be recorded on too granualar a level, and it needs to be aggregated to a higher time interval before we can draw any insight from it. Furthermore there are usually no records when there are no events. While this is logical from a data collecting point of view, it is often problematic when doing analysis like building plots or calculating a moving average. Take the following example of count data

```{r, fig.width=7}
library(ggplot2)
count <- data.frame(day =  as.Date(c('2016-07-07', '2016-07-09', '2016-07-10')),
                     cnt =  c(3,2,4))
ggplot(count, aes(day, cnt)) + geom_line() + geom_point() + expand_limits(y = 0)
```

Here we don't have an observation on the July 8 because the count here was 0. But this is missing from the data and thus from the plot. `padr` offers two functions that make light work of getting the data in the right format. `thicken` helps with aggregating the data to the desired interval, for instance summing all observations that were done on the same day. `pad` will do date padding; filling missing records so the time series becomes evenly spaced.

## Emergency data set

The package is illustrated by the `emergency` data set. It contains 120,450 emergency calls in Montgomery County, PA, between 2015-12-10 and 2016-10-17. It has four columns that contain information about the location of the emergency, a title field indicating the type of the emergency, and a time stamp. The data set was created from a Google Api, thanks to Mike Chirico for maintaining this.

```{r}
library(padr)
head(emergency)
```


## Interval

Before divinig into the two functions, we quickly discuss the meaning of interval in the `padr` context. It is the recurrence of observations in a date or time variable. It can be thought of as the heartbeat of the data, at each pulse we have an observation. The interval is "year" when we have an observation each year, "month" when there is an observation each month, etc.[^1]

[^1]: Many users that work with date and time variables will be using the `lubridate` package. The definition of an interval in `lubridate` is different from the definition in `padr`. In `lubridate` an interval is a period between two time points and has nothing to do with recurrence. Please keep this in mind.

There are eight possible values of the interval and they form a hierarchy. They are from high to low year, quarter, month, week, day, hour, minute, second.

Be aware that the interval of the data only concerns the recurrence of the data, not the value of the different time units. For instance you can have a monthly observation at any point in time within a month as long as it recurs at the exact same moment each month. A second thing to note here is that the interval definition does not require that there is an observation at each and every time point where the could be one. It is the highest recurrence level in the hierarchy that can explain all the observations. `padr` comes with the `get_interval` function to test of what interval a date time variable is.

```{r, message=FALSE}
library(dplyr)
monthly <- seq(as.Date('2016-01-01'), length.out = 12, by ='month')
get_interval(monthly)
get_interval(monthly[c(1, 3, 4, 7, 9, 11, 12)])
get_interval(emergency$time_stamp)
```

## thicken

We wouldn't be interested in the emergency data the second interval, rather we like to see developments on a daily, weekly or monthly level. This is where the `thicken` functions comes to help. As its input it takes a data frame and it looks for the date or time variable (if there are mutliple you tell it which one to use with the `by` argument). It will then add a date or datetime variable to the data frame that is of the desired interval. In combination with a framework like `dplyr` it is then quite simple to aggregate the data to this level. As an example lets make a plot of the total number of emergency calls per day.

```{r, fig.width=7, warning=FALSE}
emergency %>% thicken(interval = 'day') %>% 
  group_by(time_stamp_day) %>% 
  summarise(nr = n()) %>% 
  ggplot(aes(time_stamp_day, nr)) +
  geom_line() + geom_point()
```

Alright, what just happened here? We had a data frame containing one datetime variable (time_stamp), so `thicken` figured out which variable to use. By default `thicken` will add a column that is one interval level higher than that of the input variable. Since the interval of the input variable is second, this would have been minute. We therefore had to specify the desired interval. If not told otherwise the column name of the added variable is the name of the input datetime variable, suffixed by the interval of the thickened variable. After thickening we used `dplyr` functions to aggregate to the new interval level.

## pad

Now lets deal with the situation where there are records 'missing' from the data, since there were no observations that day. Lets look at the number of emergency calls related to dehydration. One would expect they occur more often in summertime, lets see if this is the case. First we aggregate to a daily level again.

```{r, warning=FALSE}
dehydration_day <- emergency %>% filter(title == 'EMS: DEHYDRATION') %>% 
  thicken(interval = 'day') %>% group_by(time_stamp_day) %>% 
  summarise(nr = n())
dehydration_day %>% arrange(time_stamp_day) %>% head
```

We see from the first few records that there weren't any cases of dehydration for many days in December already. We therefore need to insert records for the days obserations were missing. 

```{r, fig.width=7}
dehydration_day %>% pad %>% fill_by_value(nr) %>% 
  ggplot(aes(time_stamp_day, nr)) +
  geom_line() + geom_point() +
  geom_smooth()
  
```

Yup, definetely more dehydration in summer. So we toke the thickened time series and applied `pad` to it. The function found out that the datetime variable had a daily interval and inserted all the instances of this interval that were lacking between the first and the last observation. For all the inserted rows the `nr` variable got a missing value. The `fill_by_value` function subsequently filled all the missings. Since the default of the value argument is 0 it only needed the column to fill specified. The package provides two more fill functions. The `fill_by_function` fills all the missings by a function of the nonmissings. Often this function would be the mean or the median, but you can specify your own function if you like to. Finally there is the `fill_by_prevalent` function, that will fill the missings by the most prevalent nonmissing value. Another way that one might wish to fill the missing values is by forward filling; fill the missing by the last nonmissing value. This is already implemented in `tidyr::fill`, so there was no reaseon to redo that is this package.

## Final example

The functions in `padr` do need very little arguments to be specified when you are content with the default settings, however it allows for deviations from this. This final example shows some of this flexibility. We want to plot the weekly number of cardiac arrests. By default `thicken` let the weeks start on a Sunday, but we would like to start them on Mondays. We do so by specifying the `start_val` in `thicken`, which serves as an offset from default behavior in the intervals.

```{r}
first_day <- emergency$time_stamp %>% min %>% as.Date
first_day %>% weekdays
cardiac_week <- emergency %>% filter(title == 'EMS: CARDIAC ARREST') %>% 
  thicken(interval = 'week', start_val = first_day - 3, colname = 'weekstart') %>% 
  group_by(weekstart) %>% summarise(nr = n())
cardiac_week %>% arrange(weekstart) %>% head
```

Hmmm, thats a little strange, we have the first observations more than two months after the start of data collections. Probably a data quality issue when we looking at the number of cardiac arrests in the weeks that do have data. Lets fill these months with the median value for the purpose of illustration.

```{r, fig.width=7}
cardiac_week %>% pad(start_val = first_day - 3) %>% 
  fill_by_function(nr, fun = median) %>% 
  ggplot(aes(weekstart, nr)) + geom_bar(aes(fill = nr), stat = 'identity')
```

Ok I'll admit it, the medians at the beginning are not very insightful. However I hope you'll see from this last example that you can use the `start_val` and `end_val` argumements to extand the period of date padding beyond the period present in the data. 

## Any suggestions?
Found a bug? Ideas for improving or expandig `padr`. Your input is much appreciated. The code is maintained at https://github.com/EdwinTh/padr and you are most welcome to file and issue or do a pull request.


