---
title: "Introduction to `padr`"
author: "Edwn Thoen"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to `padr`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

When getting time series data ready for analysis, you might be confronted with the following two challenges:

* The observations are done on too low a level, e.g. time recorded to the second, where your analysis is on a daily level.
* There are no records for the time points where obsevations were absent.

`padr` aims to make light work of preparing time series data by offering the two main functions `thicken` and `pad`. A small example before we get into detail. Say I want to make a line plot of my daily expenses at the coffee place. The data for a few days might look like.

```{r, fig.width=7}
library(padr)
coffee
```

Using `padr` in combination with `dplyr` this plot is made in the following way:

```{r, fig.width=7, message = FALSE}
library(ggplot2); library(dplyr)

coffee %>% 
  thicken('day') %>% 
  group_by(time_stamp_day) %>%
  summarise(day_amount = sum(amount)) %>% 
  pad() %>% 
  fill_by_value(day_amount) %>% 
  ggplot(aes(time_stamp_day, day_amount)) + geom_line()
```

Quite some stuff going on here, let's go through the functions one by one to see what they do.

### thicken

`thicken` adds a column to a data frame that is of a higher *interval* than that of the original datetime variable. The interval in the `padr` context is the heartbeat of the data, the recurrence of the observations.[^1] The original variable "time_stamp" had the interval *second*, the added variable was of interval *day*.

```{r}
coffee2 <- coffee %>% thicken('day')
coffee2$time_stamp %>% get_interval
coffee2$time_stamp_day %>% get_interval
```

`thicken` does figure out some stuff for you. First it finds the datetime variable in your data frame (given there is only one). Next it will determine the interval of this variable, which is one of the following: year, quarter, month, week, day, hour, minute, or second. Besides the interval, it also finds the interval unit (E.g. 5 minutes, 10 days, 2 months). Finally it adds a variable to the data frame that is of a higher interval than the interval of the original datetime variable. The user can then use this variable to aggregate to the higher level, for instance using `dplyr`'s `group_by` and `summarise`. As of v.0.3.0 the user cannot only specify the interval of the desired output variable, but also the units. When no unit is specified, a single unit is applied.

```{r}
to_thicken <- data.frame(day_var = as.Date(c('2016-08-12', '2016-08-13', 
                                              '2016-08-26', '2016-08-29')))
to_thicken %>% thicken(interval = "week")
to_thicken %>% thicken(interval = "4 days")
```

We see different default behavior for the different *intervals*. Week intervals start on Sundays, day intervals start on the first day found in the datetime variable. In many situations the user will be content with `thicken`'s defaults, however you can specify the `start_val` as an offset if you would like to start the returned interval on a different day or datetime.

We use the emergency data set for further illustration. It contains 120,450 emergency calls in Montgomery County, PA, between 2015-12-10 and 2016-10-17. It has four columns that contain information about the location of the emergency, a title field indicating the type of the emergency, and a time stamp. The data set was created from a Google Api, thanks to Mike Chirico for maintaining this set. 

```{r}
head(emergency)
```

Say we are interested in the number of overdoses that occured daily. However, we don't want incidents during the same night to be split into two days, what would have happened when we use the default behavior. Rather, we reset the count at 8 am, grouping all nightly cases to the same day. The interval is still *day*, but each new day starts at 8 am instead of midnight. The `start_val` serves as an offset. 

```{r}
emergency %>% filter(title == 'EMS: OVERDOSE') %>% 
  thicken('day', 
          start_val = as.POSIXct('2015-12-11 08:00:00', tz = 'EST'),
          colname = 'daystart') %>% 
  group_by(daystart) %>% 
  summarise(nr_od = n()) %>% 
  head
```

Note also that we specified the column name of the added column. If we don't, `thicken` takes the column name of the original datetime variable and appends it with the interval of the thickened variable, separated by an underscore. 

Two final points on intervals before we are going to `pad`:

* It is about the recurrence pattern, not the moment in time where the observations toke place. (A yearly observation can take place each year at June 23rd at 9.15 pm).
* It doesn't require an observation at each and every time point where there could be one. It is the highest recurrence level in the hierarchy that can explain all the observations.

### pad

The second workhorse of `padr` is `pad`. It does date padding:

```{r}
account <- data.frame(day     = as.Date(c('2016-10-21', '2016-10-23', '2016-10-26')),
                      balance = c(304.46, 414.76, 378.98))
account %>% pad
```

The account dataframe has three observations on different days. Like `thicken` the `pad` function figured out what the datetime variable in the data frame is, and then assessed its interval. Next it noticed that within the interval, *day* in this case, rows are lacking between the first and last observation. It inserts a row in the data frame for every time point that is lacking from the data set. All non-datetime values will get missing values at the padded rows. 

It is up to the user what to do with the missing records. In the case of the balance of an account we want to carry the last observation forward. It needs `tidyr::fill` to arrive at the tidy data set.

```{r}
account %>% pad %>% tidyr::fill(balance)
```

Also `pad` allows for deviations from its default behavior. By default it pads all observations between the first and the last observation, but you can use `start_val` and `end_val` to deviate from this. You can also specify a lower interval than the one of the variable, using `pad` as the inverse of `thicken`.

```{r}
account %>% pad('hour', start_val = as.POSIXct('2016-10-20 22:00:00')) %>% head
```

Padding within groups is supported since version 0.2.0, when the `group` parameter was added to `pad`. This parameter takes one or several column names, that indicate the groups. Take the following example of the `emergency` data.

```{r, warning=FALSE}
padded_groups <- emergency %>% thicken('day') %>%
  filter(title %in% c("Traffic: VEHICLE ACCIDENT -", 
                      "Traffic: DISABLED VEHICLE -")) %>% 
  count(time_stamp_day, title) %>% 
  pad(group = 'title')
```

### fill functions
We already saw `tidyr::fill` coming in handy for the filling of missing values after padding. `padr` comes with three more fill functions: `fill_by_value`, `fill_by_function`, and `fill_by_prevalent`. They fill missing values by respectively a single value, a function of the nonmissing values, and the most prevalent value among the nonmissing values.

```{r}
counts <- data.frame(x = as.Date(c('2016-11-21', '2016-11-23',  '2016-11-24')),
                     y = c(2, 4, 4)) %>% pad

counts %>% fill_by_value()
counts %>% fill_by_value(value = 42)
counts %>% fill_by_function(fun = mean)
counts %>% fill_by_prevalent()
```
Note that in the first `fill_by_value` the columns to fill are not specified. In this case the filling is applied on all the columns. The other two functions also have this default behavior.

### One last example
`thicken` and `pad` together make a strong pair, as already was shown in the coffee example. Lets go back to the emergency data set and see how many cases of dehydration there are each day. Would there be more cases in summer?

```{r, fig.width = 7}
dehydration_day <- emergency %>% 
  filter(title == 'EMS: DEHYDRATION') %>% 
  thicken(interval = 'day') %>% 
  group_by(time_stamp_day) %>% 
  summarise(nr = n()) %>% 
  pad() %>% 
  fill_by_value(nr)
ggplot(dehydration_day, aes(time_stamp_day, nr)) +
  geom_point() +
  geom_smooth()
```

## Further details
There is a second vignette, called `padr_implementation`. Here you can find more information about how `padr` handles daylight savings time, what it does with different time zones and how `thicken` exactly is implemented.

## Any suggestions?
Found a bug? Ideas for improving or expandig `padr`. Your input is much appreciated. The code is maintained at https://github.com/EdwinTh/padr and you are most welcome to file an issue or do a pull request.


[^1]: Many users that work with date and time variables will be using the `lubridate` package. The definition of an interval in `lubridate` is different from the definition in `padr`. In `lubridate` an interval is a period between two time points and has nothing to do with recurrence. Please keep this in mind.

